---
title: "class11"
author:  "Kalle Liimatta A59002114"
format: pdf
editor: visual
toc: TRUE
---
# Section 4 : Population Scale Analysis HOMEWORK

## Q1 3 : Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

### The sample size for genotype A/A, A/G, and G/G are 108, 233, and 121, respectively.

```{r}
library(ggplot2)
```

let's make a boxplot

```{r}
box <- ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE) 
box
```

```{r}
AA <- subset(expr, geno=="A/A")
median(AA$exp)
```

```{r}
AG <- subset(expr, geno=="A/G")
median(AG$exp)
```

```{r}
GG <- subset(expr, geno=="G/G")
median(GG$exp)
```

### The median expression levels for A/A, A/G, and G/G are 31.2, 25.1, and 20.1, respectively.

## Q1 4 : Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

### Boxplot:

```{r}
box
```

### A/A is associated with significantly higher expression of ORMDL3. The SNP does affect the expression of ORMDL3.
