---
title: "class10_Halloween.Mini.Project"
author:  "Kalle Liimatta A59002114"
format: pdf
editor: visual
toc:  TRUE
---

# 1. Importing candy data

```{r}
candy_file <- read.csv("candy-data.csv")
```

```{r}
candy= data.frame(candy_file, row.names=1)
head(candy)
```

## Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

### 85 different candy types

## Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

### 38 fruity candy types

# 2. What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

## Q3. What is your favorite candy in the dataset and what is it's winpercent value?

```{r}
candy
```

```{r}
candy["Almond Joy", ]$winpercent
```

### My favorite candy in the dataset is Almond Joy. Its winpercent value is 50.35.

## Q4. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat",]$winpercent
```

### 76.77

## Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

### 49.65

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

```{r}
#alterantive way to call functions if you only need one or two functions from a package
skimr::skim(candy)
```

## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
skimcandy <- skim(candy)
skimcandy$numeric.mean
```

```{r}
skimcandy[12,]
```

### Winpercent looks to be on a different scale.

## Q7. What do you think a zero and one represent for the candy\$chocolate column?

### Zero represents that this candy is not a chocolate candy (0=no chocolate), and one represents that it is a chocolate candy (1=yes chocolate)

## Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent) + geom_histogram(bins = 10, col="red", fill="purple")
```

## Q9. Is the distribution of winpercent values symmetrical?

### No, it is slightly skewed to the left.

## Q10. Is the center of the distribution above or below 50%?

### Below

## Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
```

```{r}
mean(candy$winpercent[as.logical(candy$fruity)])
```

### On average, chocolate candy is higher ranked than fruit candy.

```{r}
chocolate.inds <-as.logical(candy$chocolate)
chocolate.win <-candy[chocolate.inds,]$winpercent
mean(chocolate.win)
```

```{r}
fruity.inds <-as.logical(candy$fruity)
fruity.win <-candy[fruity.inds,]$winpercent
mean(fruity.win)
```

## Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

### Yes, this difference is significantly different as indicated by the p-value \< 0.05.

# 3. Overall Candy Rankings

## Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

### Least liked: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

```{r}
#alternative method
inds<- order(candy$winpercent)
candy[inds,]
head(candy[inds,], n=5)
```

## Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

### Favorite candies: Reese's Peanut Butter Cup, Reese's Miniatures, Twix, Kit Kat, Snickers

## Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```

## Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col()
                  
```

```{r}
#this function will save the last plot that was made
ggsave("mybarplot.png")
#can also tell it what plot to save, how tall, the resolution, etc. (see help page for ggsave)
```

```{r}
#make all candies black as default. Then overwrite with different colors based on certain parameters.
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

## Q17. What is the worst ranked chocolate candy?

```{r}
candy %>% filter(chocolate==1) %>% arrange(winpercent)
```

### Sixlets

## Q18. What is the best ranked fruity candy?

```{r}
candy %>% filter(fruity==1) %>% arrange(desc(winpercent))
```

### Starburst

# 4. Taking a look at pricepercent

```{r}
library("ggrepel")

#how about a plot of price vs win
ggplot(candy) + aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps=5)
```

## Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

### Fruity candies are generally the best bang for your buck. Reese's miniatures is the best bang for your buck more specifically.

## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
exp <- candy %>% arrange(desc(pricepercent)) %>% head(5)
exp %>% arrange(winpercent)
```

### Most expensive: Nik L Nip, Ring pop, Nestle Smarties, Hershey's Milk, Hershey's Krackel

### Of these, Nik L Nip is the least popular.

```{r}
#alternative method
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Q21 --\> Optional

# 5 Exploring the correlation structure

```{r}
#install.packages("corrplot")
library("corrplot")
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

### Fruity and chocolate

## Q23. Similarly, what two variables are most positively correlated?

### Winpercent and chocolate

# 6. Principal Component Analysis

Need to scale our PCA plot, otherwise winpercent will dominate everything.

The main function that is always there for us is "prcomp()". It has an important argument that is set to "scale=FALSE".

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

My PCA plot (aka PC1 vs PC2 score plot)

```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
head(my_data)
```

```{r}
p <- ggplot(my_data) +
  aes(x=PC1, y=PC2,
      size=winpercent/100,
      text=rownames(my_data),
      label=rownames(my_data)) +
  geom_point(col=my_cols)
p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (pink), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

### Fruity, hard, and pluribus. These make sense as fruity candies often come in packs of multiple candies and are often hard.
